---
title: "The EPA's Environmental Quality Index (EQI) vs. Population Change"
description: |
  A statistical analysis
author:
  - name: Marie Rivers
    url: {}
date: 12-02-2021
output:
  distill::distill_article:
    self_contained: false
    code_folding: show
    toc: TRUE
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(kableExtra)
```

```{r}
library(tidyverse)
library(here)
```

This post discusses a statistical evaluation of the relationship between environmental quality and population change in the United States at the county level between 2006 and 2010.  

# Background
The Environmental Quality Index (EQI), developed by the U.S. Environmental Protection Agency (EPA) provides a county level snapshop of environmental conditions throughout the country. EPA first released EQIs for the period 2000-2005 and updated these indexes for 2006-2010. This statistical evaluation focuses on the 2006-2010 EQI. The purpose of the EQI is to use (1) as an indicator of ambient conditions/exposure in environmental health and modeling and (2) as a covariate to adjust for ambient conditions in environmental models [xxx ref].

The EQI is developed from 5 environmental domains each with identified environmental constructs as show in Table No. 1. Each county has an overall environmental index and a domain specific index. Indexes were also stratified by rural-urban continuum codes (RUCCs) for counties classified as metropolitan urbanized, non-metro urbanized, less urbanized, and thinly populated.

```{r}
eqi_domains <- data.frame(
  Domain = c("air", "water", "land", "built", "sociodemographic"),
  Constructs = c("criteria air pollutants and hazardous air pollutants", "overall water quality, general water contamination, domistic use, atmospheric deposition, drought, chemical contamination, and drinking water quality", "agriculture, pesticides, facilities, radon, and mining activity", "roads, highway/road safety, commuting behavior, housing environment, walkability, and green space ", "crime, socioeconomic, political character, and creative class representation"))
```

```{r}
kbl(eqi_domains, caption = "EQI Environmental Domains and Constructs") %>%
  kable_paper(full_width = F) %>%
  column_spec(1, bold = T, border_right = T)
```

Previous studies have used the EQI to evaluate relationships between environmental quality and public health outcomes such as:  
- preterm birth  
- mortality  
- cancer incidence  
- asthma  
- obesity  
- infant mortality

## Limitations
While the EQI can identify counties with higher environmental burdens, it may not identify environmental injustices at local community level. The EQI can not quantify environmental exposure for individuals and reflects only outside environmental conditions, not indoor conditions. The EQI can be used identify locations for future research, but is not intended for regulatory purposes or as a diagnostics tool. Due to changes in methodology and datasets, the 2000-2005 and 2006-2010 EQIs should not be directly compared. 

# The Data

## EQI
To develop the EQI, variables were identified from available data to represent each environmental domain and assessed for collinearity so redundant variables could be excluded. Then, variables were standardized based on geographic space or on a per capita rate, as appropriate. Data gaps were evaluated to distinguish between missing data and meaningful zeros. Where applicable, spatial kriging was used to interpolate values when data was not available for all counties. For example, crime likely occurs in all counties despite centralized reporting areas. Finally, the normality of each variable was assessed and transformations such as log-transformations, performed as needed.

Principal component analysis (PCA) was used to aggregate variables into domain specific indexes and domain indexes into overall indexes for each county. A result of this method was that counties with different RUCC classifications were influenced by each domain differently. Each domain does not equally influence the overall EQI value for a given county.

**Higher EQI values correspond with worse environmental quality. Lower (more negative) EQI values correspond with better environmental quality**

xxx...patchwork map of EQI (no space for maps)


## Census Population Data
County level population data was obtained from the U.S. Census Bureau's county intercensal datasets for 2000-2010. The Census Bureau estimates intercensal populations by modifying the 2000-2010 postcensal estimates using the fomula:

$$P_{t} = Q_{t}(P_{3652}/Q_{3652})^ {(t / 3652)}$$
where:  
  t = time in days since APril 1, 2000  
  $P_{t}$ = population estimate at time t  
  $Q_{t}$ = postcensal estiamte at time t  
  $P_{3652}$ = April 1, 2010 census count  
  $Q_{3652}$ = April 1, 2010 postcensal estiamte based on Census 2000  

xxx...map of population change

# Statistical Methods
A different means test was completed to determine if mean environmental quality was varied for counties experienced positive vs negative population change between 2006-2010. 

**null_hypothesis:**
There is no difference in mean EQI for counties with positive and negative population change.  

$$H_{0}: \mu_{posPopChange} - \mu_{negPopChange} = 0$$
**alternative hypothesis:** 
There is a difference in EQI for counties with positive and negative population change.

$$H_{A}: \mu_{posPopChange} - \mu_{negPopChange} \neq 0$$

```{r}
eqi_pop <- read_csv(here("_posts", "2021-11-13-the-epas-environmental-quality-index-eqi-vs-population-change", "data", "eqi_pop.csv"))

eqi_summary_table <- eqi_pop %>%
  group_by(pop_change_text) %>%
  summarise(min_eqi = round(min(EQI_22July2013, na.rm = TRUE),2),
            max_eqi = round(max(EQI_22July2013, na.rm = TRUE),2),
            mean_eqi = round(mean(EQI_22July2013, na.rm = TRUE),2),
            sd_eqi = round(sd(EQI_22July2013, na.rm = TRUE), 2),
            var_eqi = round(var(EQI_22July2013, na.rm = TRUE), 2),
            count = n()) %>%
  bind_rows(., eqi_pop %>%
              summarise(min_eqi = round(min(EQI_22July2013, na.rm = TRUE),2),
                        max_eqi = round(max(EQI_22July2013, na.rm = TRUE),2),
                        mean_eqi = round(mean(EQI_22July2013, na.rm = TRUE),2),
                        sd_eqi = round(sd(EQI_22July2013, na.rm = TRUE), 2),
                        var_eqi = round(var(EQI_22July2013, na.rm = TRUE), 2),
                        count = n()) %>%
              mutate(pop_change_text = "all counties")
            ) %>%
  kable(col.names = c("Population Change", "Minimum", "Maximum", "Mean", "Standard Deviation", "Variance", "Number of Counties"), caption = "EQI summary statistics for counties based on population change between 2006 and 2010", color = 'black') %>%
  kable_paper(full_width = FALSE) %>%
  column_spec(1, bold = T) %>%
  row_spec(0, bold = T, color = 'black') %>%
  row_spec(3, bold = T, background = '#e5e5e5')
eqi_summary_table  
```

Kalawao County, Hawaii had the lowest EQI value (highest environmental quality) and greatest decrease in population (-17.4%). Falls Church, Virginia had the highest EQI (lowest environmental quality) and a population change of 15.4% which falls is the 99th percentile.

```{r summary_statisic_values}
# mean values
mu_eqi_pos_pop_change <- eqi_pop %>% 
  filter(pop_change_pct > 0) %>% 
  summarise(mean_eqi = mean(EQI_22July2013, na.rm = TRUE))

mu_eqi_neg_pop_change <- eqi_pop %>% 
  filter(pop_change_pct <= 0) %>% 
  summarise(mean_eqi = mean(EQI_22July2013, na.rm = TRUE))

# standard deviations  
sd_eqi_pos_pop_change <- eqi_pop %>% 
  filter(pop_change_pct > 0) %>% 
  summarise(sd_eqi = sd(EQI_22July2013, na.rm = TRUE))

sd_eqi_neg_pop_change <- eqi_pop %>% 
  filter(pop_change_pct <= 0) %>% 
  summarise(sd_eqi = sd(EQI_22July2013, na.rm = TRUE))

# count of observations 
n_pos_pop_change <- eqi_pop %>% 
  filter(pop_change_pct > 0) %>% 
  summarise(count = n())

n_neg_pop_change <- eqi_pop %>% 
  filter(pop_change_pct <= 0) %>% 
  summarise(count = n())
```

```{r echo=TRUE}
# calculate point estimate
point_est <- as.numeric(mu_eqi_pos_pop_change - mu_eqi_neg_pop_change)

# calculate standard error
SE <- as.numeric(sqrt( (sd_eqi_pos_pop_change^2 / n_pos_pop_change) + (sd_eqi_neg_pop_change^2 / n_neg_pop_change) ))

# z-score
z_score <- (point_est - 0) / SE
```

$$\text{point estimate} = \mu_{posPopChange} - \mu_{negPopChange} = `r mu_eqi_pos_pop_change` - `r mu_eqi_neg_pop_change` = `r point_est`$$

The standard error for a difference in means is defined as: 

$$SE = \sqrt{\frac{s_1^2}{n_1} + \frac{s^2_2}{n_2}} = \sqrt{\frac{`r sd_eqi_pos_pop_change`^2}{`r n_pos_pop_change`} + \frac{`r sd_eqi_neg_pop_change`^2}{`r n_neg_pop_change`}} = `r SE`$$ 
and the z-score for hypothesis testing is:  

$$z = \frac{\text{point estimate - null}}{SE} = \frac{`r point_est` - 0}{`r SE`} = `r z_score`$$

# Results and Conclusion

## Future work
xxx...RUCC

While this project did not evaluate economic variables, locations with higher environmental quality could also have higher living costs which drive people to move to more affordable places. Many studies have documented to connection between environmental quality and public health. xxx[add ref]. If economic factors contribute to population growth in counties with poor environmental quality, then this could negatively affect public health in vulnerable populations and perpetuate social inequalities. Further analysis could evaluate trends in mean household income to determine if there is growing income inequality between counties with better and worse environmental quality.

In the future, EPA plans to create a census tract level EQI. This would be facilitate a higher resolution statistical evaluation of the relationship between environmental conditions and population change. 

**Data availability:**   
[EPA Datasets and files from the EQI county data from 2006-2010](https://edg.epa.gov/data/public/ORD/CPHEA/)  
[U.S. Census Bureau County Intercensal Tables: 2000-2010](https://www.census.gov/data/tables/time-series/demo/popest/intercensal-2000-2010-counties.html)

**GitHub Repo Link:**
xxx...Code used to read and tidy the data can be found here:

# References  
1. U.S. EPA. Environmental Quality Index - Technical Report (2006-2010) (Final, 2020). U.S. Environmental Protection Agency, Washington, DC, EPA/600/R-20/367, 2020.

2. U.S. Census Bureau. County Intercensal Datasets: 2000-2010. https://www.census.gov/data/datasets/time-series/demo/popest/intercensal-2000-2010-counties.html

Distill is a publication format for scientific and technical writing, native to the web.

Learn more about using Distill at <https://rstudio.github.io/distill>.


